Interactive Graphics for Visually Diagnosing Forest Classifiers in R

### 2022-06-16 PPforestVizupdate.Rnw
update plot descriptions

Figure \ref{tab2} is a screenshot of the models tab. There are five plots, with varying levels of interaction: (1) a boxplot of oob error for all trees including the data as jittered points, (2) a jittered side-by-side dotplot showing variable importance for the top three nodes of all trees in the forest, (3) a static display of one PPtree, (4) a faceted density plot of projected data at each node of the tree, with split point indicated by a vertical line, and (5) a mosaic plot showing the confusion matrix for each node of the tree.  The interaction is driven in two ways, from the boxplot plot -- when the user selects a point in that display, the corresponding importance variable plot, tree plot from the \verb#PPtreeViz#, tree density displays and mosaic plots are drawn. The tree plot from the \verb#PPtreeViz# is used to visualize the selected tree structure. Also highlighted are the variable importance values for each variable for each of the top three nodes by default and the oob error value for the tree on the boxplot.
Also it is possible other type of interaction, after selecting a tree, the user can select the nodes to show in the importance variable plot, density plot and mosaic plot (right corner).

\begin{figure}[hbpt]
\centering
\includegraphics[width=1\linewidth]{figure/tab2com.png}
\caption{The individual model tab in the web app. The oob error is displayed in a boxplot  including jittered points for all trees in the forest. Variable importance is displayed as jittered dot plots for selected nodes of all trees. The boxplot is linked to a display of the PPTree, the variable importance plot, and a density plot of the data projections showing splits at each selected nodes and confusion tables as mosaic plots. Clicking a point in the oob error boxplot triggers various updates: each of the importance values for the same tree are highlighted (red), the tree that this corresponds to is drawn, the error for the tree is shown on the boxplot (in red), and the density plot and the mosaic plot are also updated. \label{tab2}}
\end{figure}

The diagram in Figure~\ref{tab2diag} illustrates the data pipeline for interactive graphics.
The data source is a \verb#PPforest# object. Interaction is driven by the variable importance plot. Selecting a point triggers a change in the data which cascades to re-draws of the other displays. Each plot has some information available on mouse over.

\begin{figure}[hbpt]
\centering
\includegraphics[width=1\linewidth]{figure/fishT2.pdf}
\caption{Schematic diagram illustrating the interactivity in and between plots for model level exploration panel of the web app. Only the dotplot of variable importance has click selection, which invokes changes of the tree display, boxplot, density plots, and mosaic plots. Selecting a point makes a change in the data, which propagates the importance values for other variables in this tree to be highlighted (red), draws the tree, highlights the error value of the tree, and shows the projections and confusion matrix for the three top nodes.}
\label{tab2diag}
\end{figure}

@@

>>>>

Figure \ref{tab2} is a screenshot of the models tab. There are five plots, with varying levels of interaction: (1) a boxplot of oob error for all trees including the data as jittered points, (2) a jittered side-by-side dotplot showing variable importance for the top three nodes of all trees in the forest, (3) a static display of one PPtree, (4) a faceted density plot of projected data at each node of the tree, with split point indicated by a vertical line, and (5) a mosaic plot showing the confusion matrix for each node of the tree.  The interaction is driven in three ways, from the boxplot -- when the user selects a point in that display, the corresponding importance variable plot, tree plot from the \verb#PPtreeViz#, tree density displays and mosaic plots are drawn. The tree plot from the \verb#PPtreeViz# is used to visualize the selected tree structure. Also the variable importance values are highlighted for each variable for each of the top three nodes, by default, and the oob error value for the selected tree on the boxplot (both in red).
Additionally it is possible to select specific nodes and update the importance variable plot, density plot and mosaic plot (nodes can be selected on the right corner on this second tab) and finally the interaction can be driven from the variable importance plot.

The selected tree in Figure \ref{tab2} has very small oob error shown as red point in the boxplot. Two variables were use in each node partition, the first four nodes were selected and display in the importance variable plot, density plot and mosaic plot. In the importance variable plot `height` and `lenght1` were used in the first node partition being `height` the important variable to separate Bream and Parkki (Green and orange in density and mosaic plots) from the rest. In the second node `length1` and `length2` were both evenly important  to separate Bream from Parkki. A similar analysis can be done for other nodes.

\begin{figure}[hbpt]
\centering
\includegraphics[width=1\linewidth]{figure/tab2comp.png}
\caption{The individual model tab in the web app. The oob error is displayed in a boxplot  including jittered points for all trees in the forest. Variable importance is displayed as jittered dot plots for selected nodes of all trees. The boxplot is linked to a display of the PPTree, the variable importance plot, and a density plot of the data projections showing splits at each selected nodes and confusion tables as mosaic plots. Clicking a point in the oob error boxplot triggers various updates: each of the importance values for the same tree are highlighted (red), the tree that this corresponds to is drawn, the error for the tree is shown on the boxplot (in red), and the density plot and the mosaic plot are also updated. Also specific nodes can be selected to update the plots.  \label{tab2}}
\end{figure}

The diagram in Figure~\ref{tab2diag} illustrates the data pipeline for interactive graphics.
The data source is a \verb#PPforest# object. Interaction is driven in different ways, by the boxplot including jittered points. Selecting a point triggers a change in the data which cascades to re-draws of the other displays. Each plot has some information available on mouse over. Also selecting points in the variable importance triggers a change in the data and re-draws the displays corresponding to a specific tree. Finally selecting nodes update the data and re-draws the importance variable plot, density plot and mosaic plot.

\begin{figure}[hbpt]
\centering
\includegraphics[width=1\linewidth]{figure/fishT2.pdf}
\caption{Schematic diagram illustrating the interactivity in and between plots for model level exploration panel of the web app. The boxplot and the dotplot of variable importance has click selection, which invokes changes of the corresponding displays. Selecting a point in the boxplot update the node tree information (Shiny Gui) and update all the plots for the three first nodes, also nodes can be selected and update the corresponding plots. Finally selecting a point in the importance variable plot makes a change in the data, which propagates the importance values for other variables in this tree to be highlighted (red), draws the tree, highlights the error value of the tree, and shows the projections and confusion matrix for the three top nodes.}
\label{tab2diag}
\end{figure}

@@

The application have some limitations that are need to pointed out, if the number of observation is big the overploting will be an issues in some of the selectedd visualizations and if the number of variables increases, plots like parallel plot and importance variable plot should include the option to select some variables each time or a different approach.
